A free/open-source machine translation system for English to Kazakh

### 2015-03-02 engkaz-eamt-2015/engkaz-eamt-2015.tex
conversion
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59183 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\usepackage{listings}

@@

This paper describes work in progress of development of machine translation system for English-to-Kazakh, which developed by using Apertium free/open-source machine translation platform (Forcada et al. 2011, http://www.apertium.org).

\section{The Apertium platform}
Apertium (Forcada et al. 2011, http://www.apertium.org) is a free/open-source rule-based machine translation (MT) platform that was built in 2005 by the Universitat d’Alacant. At first, it was initially aimed to translate texts between closely related languages, then it was extended to deal with unrelated languages. This platform has next components: machine translation engine, developer’s tools, and linguistic data for an increasing number of language pairs and they are licensed under the free/open-source GNU General Public License (GPL, versions 2 and 3).

@@

\item 	De-formatter. Separates the text to be translated from the formatting tags.  Formatting tags are encapsulated in brackets so they are treated as “superblanks” that are placed between words in such a way that the remaining modules see them as regular blanks.  

@@

Lexical selection rules choose on of the alternative translations of one word depend on the context. This rules are written in apertium-eng-kaz.eng-kaz.lrx. Alternative translations are defined in bilingual dictionary by multiple entries of one word. For exmaple, adjective “beautiful” could be translated as “сұлу”, if following noun is subject: “beautiful girl” - “сұлу қыз”; or “әдемі”, “көркем”, if the noun means place or : “beautiful mountains” - “әдемі таулар”.  In the following table could be seen some lexical selection rules:

@@

For English–Kazakh system this stage uses three stages to improve the granularity of structural transfer rules (each one has its own rules file):
\begin{itemize}
\item A first round of transformations (“chunker”) detects source language(SL) LF patterns and generates the corresponding sequences of TL LFs grouped in chunks representing simple constituents such as noun phrases, prepositional phrases, etc. 
\item The second round (“interchunk”) reads patterns of chunks and produces a new sequence of chunks. This is the module where one can attempt to perform some longer-range reordering operations, interchunk agreement, case selection, etc. 
\item The third round (“postchunk”) transfers chunk-level tags to the lexical forms they contain and whose lexical-form-level tags are linked (through a referencing systems) to chunk-level tags (for instance, case determined for a noun phrase is transferred to the main noun), and removes all grouping information to generate the desired sequence of TL LFs.
\end{itemize}

@@

In the first round of structural transfer (the “chunker”, rules written in the apertium-eng-kaz.eng-kaz.t1x file), rules do some local operations to cut sentences into chunks, like short noun phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into Kazakh as much as possible, leaving some minor operations to be performed in later stages of structural transfer (for instance, the case of noun phrases). Table 3 describes what kind of noun phrases could appear in sentences and how it could be detected.

@@

\begin{description}
\item[Noun phrases]. Consider the following example: the chunker identifies the English sequence the large book  (determiner–adjective–noun) as a noun-phrase chunk. It translates into Kazakh, and assigns it four chunk-level tags: number (set to singular), person (set to 3rd), possessor (to be determined, as the noun кітап ('book') could receive a 3rd-person possesive ending (кітабы) later if the context were, for instance, the large book of animals, аңдардың үлкен кітабы), and case (to be determined as it could be, for instance, accusative in I saw the large book, Мен үлкен кітапты көрдім ). Also noun form of the verb, for instance,  gerund form is determined as NP phrase(Her playing is good – Оның ойыны жақсы).
\item[Prepositional phrases]. English prepositional phrase is translated into Kazakh a postpositional phrase, there are three possible outcomes with different cases:
\begin{itemize}
\item Translating the locative “-{D}{A}”,1 ablative “-{D}{A}н”, etc.,except the genitive “-{N}{I}ң” cases: [PP [P from ] [NP the five cars] ] → [PP [NP бес көлік] [P -тен ] ]
\item Genitive -NIņ, which will be marked GenP:  [PP [P of ] [NP the beautiful garden] ] → [GenP [NP әдемі бақша] [P -ның] ] 
\item 3 With noun such as аст, үст, etc. which takes possessive from main noun:  [PP [P under ] [NP the garden] ] → [PP [NP [GenP [NP бақша ][P -ның]] [NP астын]] [P -да ] ]
\end{itemize}
\item[Verb phrases]. Translation verb phrases into Kazakh  are not always be straightforward, for instance: 
present simple and future are rendered using the same tense in Kazakh (I play → Мен ойнаймын;  I will play → Мен ойнаймын), this ambiguity problem may be solved by lexical selection, which will choose right tense by watching context; 
tenses expressing continued activity, such as the English present continuous or past continuous (I am playing, I was playing), have to be detected and mapped onto sets of two lexical units (Мен ойнап жатырмын, Мен ойнап отырдым) where the main verb is found in the -п participle form (ойнап), and a suitable finite form  (жатырмын, отырдым) of an auxiliary verb (жатыр, отыр) is used to express number and person agreement1, 
modal verbs are translated by adding adjectives, which means “necessary” (керек) or “proper” (жөн), and a form of the copula (absent in present tense); the subject receives the genitive or dative case;
negative verbs for present continuous and past tense are translated by adding negative word “емес” for present perfect continuous( He has not been playing - Ол ойнап отырған емес ) and  “жоқ” for past tense and present continuous(see examples above).
\item[Translation of adjectival phrases]. In Kazakh noun phrases, adjectives come before nouns and do not show any agreement with nouns.  Adjectives can also appear in separate adjective phrases. Here are some examples:
\begin{itemize}
\item (1)  The adjective alone, marked AdjP: [AdjP big ] → [AdjP үлкен] 
\item (2)  Comparative adjective phrases  (English more + adjective, or adjective-[e]r); the Kazakh translation chooses the comparative suffix “-{I}р{A}{K}”: [AdjP  more  interesting ]  → [AdjP қызығырақ]
\item (3)  For superlative adjective phrases  “the most + adjective”  or “adjective-[e]st”, translation is built using “ең” + adjective:  [SupP  the biggest ]  → [SupP ең үлкен] [SupP  the most important ]   → [SupP ең маңызды]
\end{itemize}
Superlative adjective phrases have some properties of noun phrases (such as receiving possessive morphemes when modified by a genitive phrase: the most beautiful of people → адамдардың ең әдемісі); one could say that they are treated as NPs with an implied noun.
\end{description}

@@

The second round of structural transfer (the “interchunk” rules written in the apertium-eng-kaz.eng-kaz.t2x file) is currently performed by a proof-of-concept set of 90 rules, representative of following operations:
\begin{itemize}
\item Inter-chunk agreement between subject noun phrase and verb phrase in their person and number.
\item Assigning case to noun phrases (which are generated without case by the chunker): for instance, accusative case for objects (I bought the table → Мен үстелді  сатып алдым), genitive case for obligatory constructs (I must see → менің көруім керек),  dative case for the verb to need (I need a doctor →Маған дәрігер керек), locative case for possession (I have a book → Менде кітап бар), etc.
\item Reordering: placing of object before verb ( I[1] bought[2] the table[3] → Мен[1] үстелді[3] сатып алдым[2]),placing of prepositional pharses before the verb  (They[1] played[2] on top of the tree [3] → Олар[1] ағаштың үстінде[3] ойнады [2]), etc.
\item Adding question particle ма/ме/ба/бе in the end of question, if question mark chunk “?” detected (Did<VP\_ques> you<NP> watch<VP> last film<NP> ?<Q\_m> $\rightarrow$ Сіз<NP> соңғы фильмді<NP> көрдіңіз<VP> бе<ques> ?<Q\_m>).
\item Place possessive for long noun + noun + noun structures(The university of city of Kazakhstan - Қазақстан  қаласының университеті).
\end{itemize}
The set of rules has to be extended, as many combinations of the above phenomena are still not covered.

@@

Additional cleaning stage was created for removing extra tags which did not determine through the chunker and interchunk steps(rules are written in  apertium-eng-kaz.eng-kaz.t4x file), for instance, noun was not act as object and did not get accusative case, so not determined tag was 'CD' after “cleaning” it will be set on nominative case; at this step, also carried out agreement between noun or adjective and copula “е”(Мен дәрігер+е<p1 person singualr>), vowel and consonants agreement between last word of question and question particle(Сіз келдіңіз + ме? - Сіз келдіңіз + бе?).

@@

Current system can translate simple sentences and questions. Bilingual English-kazakh dictionary contains of  13135 entries, “chunker” rules are 168 and interchunk rules are 148. 
Evaluating was tested against  revision 58998 in the Apertium SVN1. Lexical coverage is calculated over EuroParl1, SETimes2, NewsCommentary3, Wikipedia4(last check 04/06/2014). As can be seen from  Table 5, the higher coverage is of :

\begin{table}
  \begin{tabular}{|l|r|}
    \hline
    \textbf{Corpus} & \textbf{Coverage} (\%) \\
    \hline
    Wikipedia & 84.67 \\
    EuroParl & 97.95 \\
    NewsCommentary & 96.27 \\
    SETimes & 97.90 \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{table}
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{System} & \textbf{BLEU} & \textbf{WER} \\
    \hline
    \texttt{apertium-eng-kaz} & 44.23 & 42.88 \\
    Google & 62.00 & 25.92 \\
    Sanasoft & 21.00 & 74.52 \\
    \hline
  \end{tabular}
\end{table}

@@

\section{References}
1 Сундетова А.М., Кәрібаева А.С., АПЕРТИУМ ПЛАТФОРМАСЫНДАҒЫ АҒЫЛШЫН-ҚАЗАҚ МАШИНАЛЫҚ АУДАРМАШЫ ҮШІН ЕКІТІЛДІ СӨЗДІКТІ ҚҰРУ. Матер. межд. научн.-практ. конф. «Применение информационно-коммуникационных технологий в образовании и науке», посвященной 50-летию Департамента информационно-коммуникационных технологий и 40-летию кафедры «Информационные системы» КазНУ им. аль-Фараби. 22 ноября 2013г. – Алматы: Қазақ Университеті, 2013. – С.53-57.

>>>>

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\usepackage{natbib}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\usepackage{listings}

@@

This paper describes work in progress of development of machine translation system for English-to-Kazakh, which developed by using Apertium free/open-source machine translation platform \citep{forcada11}. %(http://www.apertium.org).

\section{The Apertium platform}
Apertium \citep{forcada11} is a free/open-source rule-based machine translation (MT) platform that was built in 2005 by the Universitat d’Alacant. At first, it was initially aimed to translate texts between closely related languages, then it was extended to deal with unrelated languages. This platform has next components: machine translation engine, developer’s tools, and linguistic data for an increasing number of language pairs and they are licensed under the free/open-source GNU General Public License (GPL, versions 2 and 3).

@@

\item 	De-formatter. Separates the text to be translated from the formatting tags.  Formatting tags are encapsulated in brackets so they are treated as ``superblanks'' that are placed between words in such a way that the remaining modules see them as regular blanks.  

@@

Lexical selection rules choose on of the alternative translations of one word depend on the context. This rules are written in apertium-eng-kaz.eng-kaz.lrx. Alternative translations are defined in bilingual dictionary by multiple entries of one word. For exmaple, adjective ``beautiful'' could be translated as ``сұлу'', if following noun is subject: ``beautiful girl'' - ``сұлу қыз''; or ``әдемі'', ``көркем'', if the noun means place or : ``beautiful mountains'' - ``әдемі таулар''.  In the following table could be seen some lexical selection rules:

@@

For English–Kazakh system this stage uses three stages to improve the granularity of structural transfer rules (each one has its own rules file):
\begin{itemize}
\item A first round of transformations (``chunker'') detects source language(SL) LF patterns and generates the corresponding sequences of TL LFs grouped in chunks representing simple constituents such as noun phrases, prepositional phrases, etc. 
\item The second round (``interchunk'') reads patterns of chunks and produces a new sequence of chunks. This is the module where one can attempt to perform some longer-range reordering operations, interchunk agreement, case selection, etc. 
\item The third round (``postchunk'') transfers chunk-level tags to the lexical forms they contain and whose lexical-form-level tags are linked (through a referencing systems) to chunk-level tags (for instance, case determined for a noun phrase is transferred to the main noun), and removes all grouping information to generate the desired sequence of TL LFs.
\end{itemize}

@@

In the first round of structural transfer (the ``chunker'', rules written in the apertium-eng-kaz.eng-kaz.t1x file), rules do some local operations to cut sentences into chunks, like short noun phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into Kazakh as much as possible, leaving some minor operations to be performed in later stages of structural transfer (for instance, the case of noun phrases). Table 3 describes what kind of noun phrases could appear in sentences and how it could be detected.

@@

\begin{description}
\item[Noun phrases]. Consider the following example: the chunker identifies the English sequence the large book  (determiner–adjective–noun) as a noun-phrase chunk. It translates into Kazakh, and assigns it four chunk-level tags: number (set to singular), person (set to 3rd), possessor (to be determined, as the noun кітап ('book') could receive a 3rd-person possesive ending (кітабы) later if the context were, for instance, the large book of animals, аңдардың үлкен кітабы), and case (to be determined as it could be, for instance, accusative in I saw the large book, Мен үлкен кітапты көрдім ). Also noun form of the verb, for instance,  gerund form is determined as NP phrase(Her playing is good – Оның ойыны жақсы).
\item[Prepositional phrases]. English prepositional phrase is translated into Kazakh a postpositional phrase, there are three possible outcomes with different cases:
\begin{itemize}
\item Translating the locative ``-{D}{A}'',1 ablative ``-{D}{A}н'', etc.,except the genitive ``-{N}{I}ң'' cases: [PP [P from ] [NP the five cars] ] → [PP [NP бес көлік] [P -тен ] ]
\item Genitive -NIņ, which will be marked GenP:  [PP [P of ] [NP the beautiful garden] ] → [GenP [NP әдемі бақша] [P -ның] ] 
\item 3 With noun such as аст, үст, etc. which takes possessive from main noun:  [PP [P under ] [NP the garden] ] → [PP [NP [GenP [NP бақша ][P -ның]] [NP астын]] [P -да ] ]
\end{itemize}
\item[Verb phrases]. Translation verb phrases into Kazakh  are not always be straightforward, for instance: 
present simple and future are rendered using the same tense in Kazakh (I play → Мен ойнаймын;  I will play → Мен ойнаймын), this ambiguity problem may be solved by lexical selection, which will choose right tense by watching context; 
tenses expressing continued activity, such as the English present continuous or past continuous (I am playing, I was playing), have to be detected and mapped onto sets of two lexical units (Мен ойнап жатырмын, Мен ойнап отырдым) where the main verb is found in the -п participle form (ойнап), and a suitable finite form  (жатырмын, отырдым) of an auxiliary verb (жатыр, отыр) is used to express number and person agreement1, 
modal verbs are translated by adding adjectives, which means ``necessary'' (керек) or ``proper'' (жөн), and a form of the copula (absent in present tense); the subject receives the genitive or dative case;
negative verbs for present continuous and past tense are translated by adding negative word ``емес'' for present perfect continuous( He has not been playing - Ол ойнап отырған емес ) and  ``жоқ'' for past tense and present continuous(see examples above).
\item[Translation of adjectival phrases]. In Kazakh noun phrases, adjectives come before nouns and do not show any agreement with nouns.  Adjectives can also appear in separate adjective phrases. Here are some examples:
\begin{itemize}
\item (1)  The adjective alone, marked AdjP: [AdjP big ] → [AdjP үлкен] 
\item (2)  Comparative adjective phrases  (English more + adjective, or adjective-[e]r); the Kazakh translation chooses the comparative suffix ``-{I}р{A}{K}'': [AdjP  more  interesting ]  → [AdjP қызығырақ]
\item (3)  For superlative adjective phrases  ``the most + adjective''  or ``adjective-[e]st'', translation is built using ``ең'' + adjective:  [SupP  the biggest ]  → [SupP ең үлкен] [SupP  the most important ]   → [SupP ең маңызды]
\end{itemize}
Superlative adjective phrases have some properties of noun phrases (such as receiving possessive morphemes when modified by a genitive phrase: the most beautiful of people → адамдардың ең әдемісі); one could say that they are treated as NPs with an implied noun.
\end{description}

@@

The second round of structural transfer (the ``interchunk'' rules written in the apertium-eng-kaz.eng-kaz.t2x file) is currently performed by a proof-of-concept set of 90 rules, representative of following operations:
\begin{itemize}
\item Inter-chunk agreement between subject noun phrase and verb phrase in their person and number.
\item Assigning case to noun phrases (which are generated without case by the chunker): for instance, accusative case for objects (I bought the table → Мен үстелді  сатып алдым), genitive case for obligatory constructs (I must see → менің көруім керек),  dative case for the verb to need (I need a doctor →Маған дәрігер керек), locative case for possession (I have a book → Менде кітап бар), etc.
\item Reordering: placing of object before verb ( I[1] bought[2] the table[3] → Мен[1] үстелді[3] сатып алдым[2]),placing of prepositional pharses before the verb  (They[1] played[2] on top of the tree [3] → Олар[1] ағаштың үстінде[3] ойнады [2]), etc.
\item Adding question particle ма/ме/ба/бе in the end of question, if question mark chunk ``?'' detected (Did<VP\_ques> you<NP> watch<VP> last film<NP> ?<Q\_m> $\rightarrow$ Сіз<NP> соңғы фильмді<NP> көрдіңіз<VP> бе<ques> ?<Q\_m>).
\item Place possessive for long noun + noun + noun structures(The university of city of Kazakhstan - Қазақстан  қаласының университеті).
\end{itemize}
The set of rules has to be extended, as many combinations of the above phenomena are still not covered.

@@

Additional cleaning stage was created for removing extra tags which did not determine through the chunker and interchunk steps(rules are written in  apertium-eng-kaz.eng-kaz.t4x file), for instance, noun was not act as object and did not get accusative case, so not determined tag was 'CD' after ``cleaning'' it will be set on nominative case; at this step, also carried out agreement between noun or adjective and copula ``е''(Мен дәрігер+е<p1 person singualr>), vowel and consonants agreement between last word of question and question particle(Сіз келдіңіз + ме? - Сіз келдіңіз + бе?).

@@

Current system can translate simple sentences and questions. Bilingual English-kazakh dictionary contains of  13135 entries, ``chunker'' rules are 168 and interchunk rules are 148. 
Evaluating was tested against  revision 58998 in the Apertium SVN1. Lexical coverage is calculated over EuroParl1, SETimes2, NewsCommentary3, Wikipedia4(last check 04/06/2014). As can be seen from  Table 5, the higher coverage is of :

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) \\
    \hline
    SETimes & 5.1 million & 97.90 \\
    NewsCommentary & 6.5 million & 96.27 \\
    EuroParl & 54.5 million & 97.95 \\
    Wikipedia & 1.83 billion & 84.67 \\
    \hline
  \end{tabular}
  \caption{Vocabulary coverage of the English--Kazakh system over four available corpora.}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{System} & \textbf{BLEU} & \textbf{WER} \\
    \hline
    \texttt{apertium-eng-kaz} & 44.23 & 42.88 \\
    Google & 62.00 & 25.92 \\
    Sanasoft & 21.00 & 74.52 \\
    \hline
  \end{tabular}
  \caption{Metric results for the three systems compared.}
  \label{table:metrics}
\end{table}

@@

%\section{References}
%1 Сундетова А.М., Кәрібаева А.С., АПЕРТИУМ ПЛАТФОРМАСЫНДАҒЫ АҒЫЛШЫН-ҚАЗАҚ МАШИНАЛЫҚ АУДАРМАШЫ ҮШІН ЕКІТІЛДІ СӨЗДІКТІ ҚҰРУ. Матер. межд. научн.-практ. конф. «Применение информационно-коммуникационных технологий в образовании и науке», посвященной 50-летию Департамента информационно-коммуникационных технологий и 40-летию кафедры «Информационные системы» КазНУ им. аль-Фараби. 22 ноября 2013г. – Алматы: Қазақ Университеті, 2013. – С.53-57.

\bibliographystyle{apalike}

\bibliography{2015-eamt-engkaz}
